\documentclass{thiguka}
\usepackage{bookmark}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{lmodern}
\usepackage{multicol}
\usepackage[margin=4em]{geometry}

\usepackage{gb4e}
\setlength{\glossglue}{5pt plus 2pt minus 1pt}

\usepackage{array}
\usepackage{tipa}
\usepackage{textcomp}

\usepackage{fontspec}
\setmainfont{Gentium Plus}

\usepackage{booktabs}
\usepackage{tabularx}

\usepackage{iftex}
\iftutex
  \DeclareFontFamilySubstitution{T3}{\rmdefault}{cmr}
\fi

\begin{document}

Ambagame3 Translation

From: jjommoma

To: Lemuria

June 16, 2024

\section{Translation}
\subsection{Thiguka}
Fipelupahelah falusaya geguelah geylepahelah.
Isa geylepafaysa leylepahsa lasa “Sutepafay-rukahba.”\\
Sutepafay-rukahbakala, pelupah sutepahelah suguiba sutepafayelah ibilayelah.
Geylepafaypa sifarlarikigalpahsa lasa rakuka sigutu sifarabukakuspafay sutepafayelah ibilaktay.
Pelupah baseba itala sutetayelah.
Geylepahsa sakara kula isa giledpahsa pafay sutetaypa.

\subsection{Lemuria's translation}

\subsection{Gloss}

\begin{exe}
\ex{} \gll{}Fi-pelu-pah-elah falusaya ge~gu-elah geyle-pah-elah.\\
            DIM-person-NOM-PL like AGR~ADJ-many game-NOM-PL.\\
      \glt{}Children like a lot of games.
\end{exe}

\begin{exe}
\ex{} \gll{}Isa geylepafaysa leylepahsa lasa "Sute-pafay---rukahba."\\
            one game-GEN-SG name-NOM-SG COP.1SG.PRS stone-GEN-fight.LOAN.V\\
      \glt{}One of these games is named "Stone Fighting".
\end{exe}

\begin{exe}
\ex{} \gll{}Sute-pafay---rukahba-kala, pelu-pah sute-pah-elah su~gu-iba sute-pafay-elah ibi-lay-elah.\\
            stone-GEN-fight.LOAN-LOC, person-NOM stone-NOM-PL AGR~ADJ-other stone-GEN-PL stack.LOAN-DAT-PL\\
      \glt{}In Stone Fighting, people throw stones at other stacks of stones.
\end{exe}

\begin{exe}
\ex{} \gll{}Geyle-pafay-pa sifarlarikigal-pah-sa lasa rakuka su~gu-tu sifarabukakus-pafay sutepafayelah ibilak-tay.\\
            game-GEN-ZERO objective.LOAN-NOM-PL COP.1SG.PRS topple AGR~ADJ-all opponent-GEN stone-GEN-PL stack.LOAN-ACC\\
      \glt{}The game's objective is to topple the opponent's every stack of stones.
\end{exe}

\begin{exe}
\ex{} \gll{}Pelu-pah baseba itala sute-tay-elah.\\
            person-NOM can.LOAN catch.LOAN stone-ACC-PL\\
      \glt{}People can catch stones.
\end{exe}

\begin{exe}
\ex{} \gll{}Geyle-pah-sa sakara kula isa giled-pah-sa pafay sute-tay-pa.\\
            game-NOM-SG stop if one group-NOM-SG has stone-ACC-ZERO\\
      \glt{}The game stops if one group has no stones left.
\end{exe}

\section{Commentary}
\begin{enumerate}
    \item The long dash --- indicates dashes in original source text that are supposed to be there.
    \item *''geylepahelah'' should be marked as ''geyletayelah'', since it is accusative; being the object.
    \item I interpreted ''ibilayelah'' as the dative plural form of ''ibilak'' ("stack, pile").
    \item I am confused by "Geyle-pafay-pa"; it means "no game", but it may also be a misspelling of "pah".
    \item Using the 1SG.PRS copula ''lasa'' is incorrect here; you are not the game, you are talking about the game in the third person perspective. I would have used ''gulaya'' here.
\end{enumerate}

\newpage

\section{New coins}
\begin{enumerate}
    \item rukahba, ``to fight'', from ruqawūd
    \item ibilak ``tack, pile'', from ibilak
    \item sifarlarikigal ``oal, objective'', from şifarnarikigan
    \item rakuka, ``to topple'', from raqykad
    \item sifarabukakus, ``opponent'', from şifarabukakyz
    \item baseba, ``to be able to; can'', from majebad
    \item itala, ``to catch'', from yetanud
    \item tu, ``all'', from tu
\end{enumerate}

\end{document}
