\documentclass{thiguka}
\usepackage{bookmark}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{lmodern}
\usepackage{multicol}
\usepackage[margin=4em]{geometry}

\usepackage{array}
\usepackage{tipa}
\usepackage{textcomp}

\usepackage{fontspec}
\setmainfont{Gentium Plus}

\usepackage{booktabs}
\usepackage{tabularx}

\usepackage{expex}

\begin{document}

Ambagame3 Translation

From: Zethar

To: Lemuria

June 28, 2024


\section{Translation}

\subsection{Original}
Kakepah fusala rikaritifeylatha, thakapah safathusisa ithesipakalataysa pelukithisa leyfatapah sagulaya gaguliasaga gagulitelosolila gasaritayelah:

\begin{enumerate}
\item A: Kakepah gesalahsa.
\item B: Kakepah fusala karigulid salahsatay rilahpeledpahelah sagisafi kisu thakatay li fathaslay.
\item A: Kakepah pagasari saka paslakala saleyfatapafay alu lisurepafay kalisu?
\item B: Taguisa talit asila pelutayelah. Kafugudalata thakapah kasgasari alu filayfulafata.
\end{enumerate}

\subsection{English}
You need to listen to me. A hacker and I are hacking into the comms network right now and listening in on calls.

\begin{enumerate}
\item A: You're isolated.
\item B: You better hide. Those corporations want me to disappear and die.
\item A: You act like working there is fun. Why so?
\item B: I'm trying to help someone. I guess I might go hike with my cat.
\end{enumerate}

\section{Gloss}
Glosses below are literal.

\ex
\begingl
\gla  Kakepah fusala ri-ka-ritifeyla-tha//
\glb  2SG-NOM must.be INT-INTEN-listen-FUT//
\glft ``You must be intently listening.''//
\endgl
\xe

\ex
\begingl
\gla thakapah sa-fathusi-sa ithesipakala-tay-sa pelu-kithi-sa leyfata-pah sa-gulaya ga\~{}gu-li-asaga ga\~{}gu-li-telosolila gasari-tay-elah.//
\glb 1SG-NOM IPFV-expose.V-PRS internet.voice.chat-ACC-SG person-INS-SG career-NOM IPFV-COP.3SG.PRS AGR\~{}ADJ-NEG-allow AGR\~{}ADJ-NEG-technology action-ACC-PL//
\glft A hacker and I are hacking into the comms network right now and listening in on calls.\\
      lit. I'm exposing internet voice chats with someone whose career is not allowing technology to act.//
\endgl
\xe

\ex
\begingl
\gla A: Kake-pah ge-salahsa.//
\glb A: 2SG-NOM PROG-remote//
\glft lit. You're isolated.//
\endgl
\xe

\ex
\begingl
\gla B: Kakepah fusala ka-ri-gulid salahsa-tay ri-lahpeled-pah-elah sa-gisafi kisu thaka-tay li fathas-lay.//
\glb B: 2SG-NOM must INTEN-INT-good remote-ACC INT-company-NOM-PL IPFV-desire kill 1SG-ACC NEG see-DAT//
\glft lit. You better be well-isolated. Those corporations want to kill me and make me disappear.//
\endgl
\xe

\ex
\begingl
\gla A: Kake-pah pa-gasari saka pasla-kala saleyfata-pafay alu lisure-pafay kalisu?//
\glb A: 2SG-NOM PFV-act MED.DET place-LOC occupation-GEN and fun.activity-GEN what?//
\glft You act as if working there is fun. Why?
      lit. You act like that place's occupation is fun. What?//
\endgl
\xe

I have exclusively used the genitive case to indicate posession in Thiguka, so Zethar using it for something else threw me off.
Clearly, it's another "bad documentation" problem.
All I can say to this sentence is... \textit{kalisu}?

\ex
\begingl
\gla B: Ta\~{}gu-isa talit asila pelu-tay-elah. Kafu~gu-dalata thakapah kas-gasari alu filay-fulafata.//
\glb B: AGR\~{}ADJ-one help DIST.DET person-ACC-PL ADJ-two 1SG-NOM CONTEM-act and with.cat-hike//
\glft That person could use some help. I've gotta act and go hike with my cat.\\
      lit. I'd help one person. I guess I might go hike and bring my cat.//
\endgl
\xe

I am confused by the meaning of \textit{Kafu} in this sentence.
It made me go \textit{kalisu} yet again, and unfortunately because I couldn't stop myself from procrastinating, it's hopeless.


\section{New coins}
\begin{enumerate}
    \item talit: to help (from Cowlitz talíč- [taˈlitʃ-] “help”)
\end{enumerate}

\section{Miscellaneous}

From Zethar:

\begin{quote}
The text is a little short because the grammar and the lexicon doesn’t mesh well with the conculture in my conlang, so I have to write something slightly different.
I also needed to prep for Shadowrun so for background I’m imagining this take place in a dystopian world where megacorporations control everything and has outlawed magic which has been re-surging because they don’t control it.
I focused mostly on exercising the grammar which I was given, and I think I have pretty good coverage, at least containing something from each section.

I think the translation would be fairly difficult mostly because of the under-documentation and guessing I had to do, as well as the circumlouctions I employed to avoid coining things.

Good luck, Zethar
\end{quote}


\end{document}
